---
norm---
title: "bombus_life_history"
format: html
editor: visual
---

## Bombus spp. life history model

We are creating a R model for Bombus spp. that provides estimates of annual colony success. Requirements include that the model incorporates the effects of daily exposures to time-varying chemical contamination in a range of environmental medias (surface water, soil, pollen, nectar, air) and accepts seasonally changing foraging suitability maps. A population consists of an initial queen and daily cohorts for larvae, workers, and drones.

The implementation is similar to BeePop+ (DGH, Garber) and Banks et al (2020), with a daily difference equation approach to feasibly incorporate daily forcing variables that include exposure media concentrations; weather variables that affect foraging and thermoregulation; and spatially defined foraging quality. Exposures for Bombus spp. differ due to consumption rates, life history parameters, and chemical concentrations in nesting materials (soil, leaves, wax). Three Bombus species are parameterized (B. affinis, B. impatiens, B. terrestris). Multiple metrics for evaluating annual colony viability are considered.

## Life history timeline parameters

For simplicity, we will specify deterministic queen emergence and other major life history transition dates within a Bombus colony season. However, it is important to note that degree-day accumulation approaches are strong options and also important for determining intersections between pesticide exposures and important Bombus life history events due to their utility in estimating crop phenology and pest insect phenomena.

We begin with a presumed successful emergence of a queen from the prior year (not connected to prior year's overwintering success). She must then establish a nesting site in a suitable location, typically dense grassland or an underground burrow. The queen then starts laying eggs at a species-defined rate, these eggs develop into the first daily cohorts of worker bees while also foraging for herself and larval resources. After maturation of the first worker cohort, the female workers take over pollen/nectar foraging, brood/larval care, and nest expansion duties.

Although requeening events are possible, we do not consider them. Initial competition for nest locations from competing emergent queens does occur and late emergent queens can steal colonies from existing queens or take over queenless colonies. These event probabilities are a function of emergent queen densities and available habitat (Darvill et al 2004; Knight et al 2005; Najamitsu and Yamagiahi 2009; OConnor et al 2012; Osborne et al 2007).

```{r}
#| echo: false
library(gt)
df_life_history_events <- data.frame(
    Parameters = c("Queen emergence", "Initial nest construction", "Worker egg laying initiation", 
                   "Gyne/male egg laying initiation", "End egg laying"),
    "B affinis" = c("", "", "", "", ""),
    "B impatiens" = c("", "", "", "", ""),
    "B terrestris" = c("", "", "", "", "")
)
table_life_history_events <- gt(df_life_history_events)
table_life_history_events |> as_raw_html()
```

## Bombus traits

It will be necessary to borrow parameters from other Bombus species. B. affinis is part of the subgenus Bombus, therefore priority will be given to other species within this same subgenus, principally B. occidentalis (Western bumble bee), B. terricola (Yellow-banded bumble bee). Other species include cryptarum, franklini, ignitus, lucorum, and magnus. For toxicity data we will rely primarily on Bombus terrestris and Bombus impatiens.

## Bombus life history

Bombus classes.

### Foundress

The simulation begins on January 1 (calendar day 1) with a hibernating queen. She emerges from hibernation on a specified day (April 10th) and begins to forage and search for a nesting site. She is exposed to pollen, nectar and air concentrations as she forages and soil concentrations during nest building activities.

She is assumed to emerge on a day given by n(100,10), rounded (McVicker 2023, Szymanski et al 2016); though Prys-Jones gives N(91,28) for B. terrestris.

```{r}
#| echo: false

num_simulations <- 1000
foundress_emergence_date <- rnorm(num_simulations, mean = 100, sd = 10)
```

We can define the successful establishment of a nest by the production of brood. After emerging, she is assumed to have a \~60% chance of successfully establishing a nest (B. occidentalis 119/201 = 59.2% nest initiation success rate, Strange et al. 2023) in the absence of any external stressors. If she does not establish a nest by 21 days after emergence then it is assumed that she has failed. Therefore we can estimate a daily probability of nest initiation success over a 21 day period that approximates the observed nest initiation failure rate of 40.8%. This can be derived by specifying a daily nest initiation success rate of 4%, giving an overall probability of failure given by (1-0.04)\^21 = 42.4% after 21 days, which is reasonably close to 40.8% from Strange et al. (2023). We also precede this calculation with an initial 4 day post-emergent period with a 0% daily chance of nest initiation to give the foundress time to forage and to better match the average time to first brood given by Strange et al (2023) of 13.7 with a standard deviation of 9.2 days.

Simulate time to nest initiation for 1000 foundresses.

```{r}
#| echo: false
# Function to simulate time to success over multiple days
simulate_time_to_success <- function(p) {
  days <- 0
  success <- FALSE
  while(!success) {
    days <- days + 1
    success <- runif(1) < p
  }
  return(days)
}

# Simulate time to success over multiple days
p <- 0.04
nest_initiation_results <- replicate(num_simulations, simulate_time_to_success(p))
nest_failure_rate = length(nest_initiation_results[which(nest_initiation_results>21)])/num_simulations
nest_initiation_results[which(nest_initiation_results>21)] <- NA
nest_initiation_results <- nest_initiation_results + 4
mean_initiation_time <- mean(nest_initiation_results, na.rm=T)
sd_initiation_time <- sd(nest_initiation_results, na.rm=T)
# Visualize the distribution of time to success
hist(nest_initiation_results, xlab = "Days to nest initiation", main = paste0("N=", num_simulations, "; failure rate = ", nest_failure_rate, 
                                                                              "; initiation time = ", round(mean_initiation_time,2), " +/- ", 
                                                                              round(sd_initiation_time,2), " days"))
```

We can further define the establishment of a nest with the successful production of at least one worker from the brood, at this point the foundress becomes a queen. From Strange et al. (2023), of the 119/201 foundresses who successfully initiated a nest, 70 of them successfully produced workers. Giving a conditional establishment success rate of 58.8% (34.8% overall). From these values, we can see that each queen needs to produce more than 3 fertilized gynes at the end of the season to maintain population numbers over multiple years.

Strange et al. (2023) also observed 35.0 +/- 7.1 days to the emergence of a worker, an average of 21.3 days post nest initiation and associated brood production. We therefore apply a binomial probability to each successfully initiated nest to determine if workers are produced and then sample from a rounded N(21.3,3) time variable to give the discrete number of days after initiation for nests that are successfully established.

```{r}
#| echo: false
nest_establishment_results <- rnorm(num_simulations, mean = 21.3, sd = 4)
worker_emergence_results <- nest_initiation_results + nest_establishment_results
nest_establishment_failure_rate <- 0.412
nest_establishment_failures <- rbinom(num_simulations, 1, (1-nest_establishment_failure_rate))
worker_emergence_results <- worker_emergence_results * nest_establishment_failures
worker_emergence_results[which(worker_emergence_results==0)] <- NA
mean(worker_emergence_results, na.rm=T)
sd(worker_emergence_results, na.rm=T)
```

Colony size: 250-1341 bees (Boone et al. 2022)

Average maximum flight distance: 991m (Boone et al. 2022)

Day emerging: April 10th (100, sd:10) (Symanski et al, 2016)

Day of hibernation: October 10th (283, sd:10) (Symanski et al, 2016)

Days Queen active: April 10th-May 31st (Goulson et al. 2003)

Day Queen establishes nest and does not leave: May 31st? (Symanski et al, 2016)

## Load media concentrations

## Exposure calculations

We need to translate the media concentrations brought back by the queen and/or encountered in the landscape to tangible exposure metrics. We want estimates for toxicity via oral ingestion of nectar and pollen, as well as contact from soil and depsoition. We have to consider queen ingestion rates per day depending on timing, as well as per day soil contact exposure. Results for approximate surface area come from worker size from Kuhsel et al. 2017, and ingestion rates for workers are from Gradish et al. 2019. Daily foraging trip and vertical dispersal heights come from Krupke et al. 2017. Estimates for surface area are adapted by increasing the supposed surface area by the average increase in size between worker and queen bumblebee).

***CONTACT:***

**\--Aerial Deposition Contact:**

Krupke et al. 2017: 'the final estimate for the flight time portion of a day of foraging was the product of the proportion of planting time spent in 10 round trips across a field and surrounding zones, the volume of air encountered by a forager per square meter the distance of 10 round trips, and the concentration for a given dust cloud height." We will negate the specific field area, and instead, assume that the bee has the potential to fly through the dust or drift cloud across the entire distance in a single foraging trip, accounting for a bumblebee specific 'flight tube' representing the volume of air passed through per m. We will do a conservative estimate, and assume that the dust cloud is distributed

To calculate the area of the 'flight tube', we adjusted the approach used by krupke et al. (2017).

The flight tube represents the hypothetical volume of 'air' that a forager passes through in flights to and from colony. We calculate it by taking an approximate diameter and horizontal surface area that includes the wing length.

HW: 9mm (bombus affinis queen, from Xerces)

Wing surface area: 197mm2 (bombus terrestris, Nagamitsu et al. 2006)

As follows similarly in Krupke et al. (2017), we calculate the volume of air passed through per m as the diameter of a circle representing the head width, plus half of the total wing area, across 1 meter of flight.

Wing area, in mm: 197mm2/2 = 98.5mm2 (divide by two for a single side)

98.5mm2 + pi(4.5 mm)\^2 = area of cross-sectional cylinder

area of cross-sectional cylinder converted to meters2 \* 1m length = volume (m3) / per meter distance traveled

We have the distance traveled in a single foraging bee estimated as 991m per day (per Boone et al. 2022, close to the estimate of 82 mins and 5m/s flight speed, assuming a more resource scarce environment from Westphal et al. 2006, in line with noted behavior of staying close to nest observed in RPBB)

We have the deposition distributed throguh 2m height as Deposition X (ug/m2)/2m dispersal height = Deposition X (ug/m3)

We calculate the ug/bee in a day as:

**Individual Drift Contact (ug/bee) = Deposition (ug/m3) \* Volume of flight tube (1.6 E-4 m3) / m \* Distance traveled (991 m/bee )**

**\--Flower Deposition Contact:**

Krupke et al. assume that the deposition on dandelions is relative to the deposition off-field per cm2

We will calculate the exposure to per plant deposition simply as the repeat exposure to deposition off-field, some n number of times. We use the value from Heinrich 1976 for Bombus impatiens, where an individual foraging bee visited approximately 15 flowers per trip, with 6 trips, i.e., 90 repeat events (Heinrich 1976).

**Individual Flower Contact (ug/bee) = Deposition (ug/m2) \* *0.0001* \* 6.5cm2 (surface area of individual bee) \* 90**

**\--Soil Contact:**

5cm2 surface area of bumblebee \~+ 1.5 (based on average increase in size from worker to queen length)

Can we use soil depth and bulk density to estimate a ug/cm2? and multiple that times the surface area? This is basically back calculating the way we converted the ug/g to that concentration in the first place,

X ug/g soil concentration x 2 cm depth of soil spread x 1.95 g/cm3 x 5cm2/bee = X ug/bee contact?

***INGESTION:***

\--FORAGING WORKER: 0.400g/bee/day for nectar

\--FORAGING WORKER: 0.030g/bee/day for pollen

## Foraging

Accounting for foraging days based on weather.

## Toxicity data

Contact LD50 for Clothiandin to honey bees is 43.9 ng/bee (EPA 2003). Oral LD50 of 25.4 ng/bee (Yao et al. 2018).

From the RPBB risk assessment/lit review/general:

```{r}
#| echo: false
library(gt)
table_params<-read.csv(paste0(root_data_in,"/PesticideData/BeeTox.csv"))
table_params <- gt(table_params)
table_params |> as_raw_html()
```

## Converting hazard ratios to mortality probabilities

Normalized to 24 hours.

## Results
